Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC and DLPFC with TMM normlaisation and Custom Design 1)

### 2016-11-30 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_CD1.Rmd
Added more technical variables to the custom design matrix

cl = makeCluster(14)
registerDoParallel(cl)

@@

```{r dataExplore, fig.width = 6, fig.height = 15}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

multiplot(plotlist = p, cols = 1)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

### Normalisation (custom design)
Normalising COUNTS with a custom design matrix. Here Instituiton, Gender, LibraryBatch, Age\\_of\\_Death, PMI, RIN, RIN2, EV.1, EV.2, EV.3, EV.4, EV.5 are adjusted conditioned on Dx.Tissue with Individual\\_ID as random effects.
```{r iterativeNorm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'LibraryBatch', 'Age\\_of\\_Death',
                     'PMI', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5');

@@

# Residual covariates to choose from
residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Individual_ID'))

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES[, residCovars, drop=F], 
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```

@@

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tissue) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tissue) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

@@

@@

# Store design matrix
write.table(DM1$design, file = 'CMC_ACC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_ACC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

>>>>

cl = makeCluster(38)
registerDoParallel(cl)

@@

```{r dataExplore, fig.width = 5, fig.height = 5}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1),
                  plot.title=element_text(hjust=0.5))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
p[[1]]

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme
p[[2]]

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme
p[[3]]

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme
p[[4]]

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme
p[[5]]

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TranscriptsDetected)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Transcripts Detected') + my.theme
p[[6]]

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme
p[[7]]

# PercentAligned
p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PercentAligned)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Percent Aligned') + my.theme
p[[8]]

# rRNARate
p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme
p[[9]]

# Institution
vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)

# Gender
vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

### Normalisation (custom design)
```{r iterativeNorm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'IntronicRate', 'IntergenicRate', 'MappedReads', 'PercentAligned',
                     'RIN', 'PMI', 'Age_of_Death', 'rRNARate', 'RibozeroBatch', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5',
                     'GenesDetected');

@@

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES[, ind.cov, drop=F], 
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "transcript_length"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GC.CONT = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
  unique
rownames(GC.CONT) = GC.CONT$ensembl_gene_id

Gx.LEN = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, transcript_length) %>%
  group_by(ensembl_gene_id) %>%
  summarise(transcript_length = max(transcript_length, na.rm=T)) %>%
  unique() %>% as.data.frame()
rownames(Gx.LEN) = Gx.LEN$ensembl_gene_id

# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
              unique()) %>%
  unique

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
    left_join(Ensemble2HGNC %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
              unique()) %>%
  unique
```

@@

tmp = DE %>%
  dplyr::mutate(dir = logFC/abs(logFC)) %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue, dir) %>%
  group_by(Comparison, Tissue, dir) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::mutate(dir = logFC/abs(logFC)) %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue, dir) %>%
  group_by(Comparison, Tissue, dir) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

@@

```{r gc.gx, fig.width = 7, fig.height = 14}
tmp = left_join(DE, GC.CONT) %>%
  left_join(Gx.LEN) %>%
  unique

p = list()
p[[1]] = ggplot(tmp, aes(x = log10(transcript_length), y = logFC)) + geom_point() + geom_smooth(method = 'lm')
p[[1]] = p[[1]] + xlab('Log10 of Maximum Tx Length')

p[[2]] = ggplot(tmp, aes(x = log10(percentage_gc_content), y = logFC)) + geom_point() + geom_smooth(method = 'lm')
p[[2]] = p[[2]] + xlab('Percent GC Content')
CovariateAnalysis::multiplot(plotlist = p, cols = 2)
```

@@

# Store design matrix
write.table(DM1$design, file = 'CMC_ACC_Design.tsv', sep = '\t', row.names=T, quote=F)
DM_OBJ = File('CMC_ACC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)
